---
title: "Kimura 2-parameter model"
author: "Ben Nolan"
date: "4/18/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Using the Kimura 2-parameter model to calculate maximum-likelihood for genetic distance

```{r, include=FALSE}
#load in the libraries
library(expm)
library(MASS)
library(optimx)
library(seqinr)
```


For the kimura 2-parameter model, the `kappa` (alpha:beta ratio) and `t` (time) parameters are optimised from the maximum value of genetic distance from two sequences, using optimx. `t` is scaled to be equal to 1 mutation per unit time, allowing the calculation of genetic distance.

```{r}
#Function to convert nucleotides to numbers
s2n <- function(seq){
  num_seq = c()
  nucs <- c("A", "C", "G", "T")
  #split into vector
  seq_split = unlist(strsplit(seq, ""))
  #convert to numbers 1:4
  for(i in 1:length(seq_split)){
      num_seq[i] <- match(seq_split[i], nucs)
  }
  return(num_seq)
}
```

```{r}
kimura2 <- function(pars, seq_a, seq_b){
  #kappa=alpha/beta ratio
  #t=time interval
  k <- pars[1]
  t <- pars[2]
  
  #alpha and beta
  a = k/(2+k)
  b = 1/(2+k)
  
  #calculate  Generator matrix 
  G = matrix(c(-2*b-a, b, a, b,
               b, -2*b-a, b, a,
               a, b, -2*b-a, b,
               b, a, b, -2*b-a),
             c(4,4))
  
  #scale G so that the sum of every off diagonal position in each row equals 1
  #pi=1/4 under kimura 2-parameter; 1/4*(4*a+8*b) = 2*b + a
  G_scaled = G / (2*b + a)

  #calculate TPM
  P = expm(t * G_scaled)
   #calculate probability of any aligned pair of sequences
   #input sequences converted to numbers using s2n function
   seq_a = s2n(seq_a)
   seq_b = s2n(seq_b)
   
   length = length(seq_a)
  
   #calculate the genetic distance
   #create variable
   prob = 1
   #add probabilities for any two sequences
   for(i in 1:length){
     prob = prob * P[seq_a[i], seq_b[i]]
   }

  return(prob)
}

```

### Example

Obtain a maximum likelihood estimate of the genetic distance of the following two sequences using the kimura 2-parameter model:

```
A G T C C A T G A T
A C G T C G T G C T
```

```{r}
#maximum likelihood estimate of the k (kappa, or alpha/beta ratio) 
#and t (time which is scaled to be equivalent to genetic distance)

seq_a = "AGTCCATGAT"
seq_b = "ACGTCGTGCT"

#Using optimx to find the maximum likelihood 
optimx(c(k=1, t=1), kimura2, seq_a=seq_a, seq_b=seq_b, control = list(maximize=T))
```


Here, `t` is scaled to be 1 mutational change per unit of time. `t` is equivalent to genetic distance. In this particular test of:

```
A G T C C A T G A T
A C G T C G T G C T
```

The genetic distance is 0.83 under the Nelder-Mead model, or 0.822 under the BFGS. 

* The maximum likelihood estimate of the genetic distance (Nelder-Mead), given the above pair of sequences is:
  + **Probability of observed data:** 4.218750e-06
  + **t:** 0.83
  + **k:** 1.63.

